---
title: "enviPathR"
author:
  - name: Giulio Benedetti
    affiliation: University of Turku
    email: giulio.benedetti@utu.fi
output:
    BiocStyle::html_document:
        fig_height: 7
        fig_width: 10
        toc: true
        toc_float: true
        toc_depth: 2
        number_sections: true
        self_contained: true
        code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('enviPathR')`"
vignette: >
  %\VignetteIndexEntry{enviPathR: access enviPath and predict pathways from R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

username <- Sys.getenv("EP_USERNAME")
password <- Sys.getenv("EP_PASSWORD")
```

## Introduction

enviPath is a biotransformation database and pathway prediction system focusing
on xenobiotic compounds that exacerbate environmental pollution. Although the
enviPath resources are accessible via an intuitive website, this approach cannot
ensure analytical efficiency and reproducibility. For this, enviPath provides
a REST API through which both data retrieval and pathway prediction are
possible. The current package, enviPathR, offers an R client to the enviPath API
and thus brings most of its capabilities to R practitioners studying
environmental pollution and organic contaminant biotransformation.

## Tutorial

This section outlines how to use enviPathR for several different purposes. While
enabling all resource look-up and pathway prediction functionality, the package
currently lacks the API methods for object creation and deletion. That can be
added upon request (external contributions are also welcome).

### Installation

Bioconductor release version:

```{r install-release, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("enviPathR")
```

Beta version:

```{r install-devel, eval = FALSE}
remotes::install_github("Minotau-R/enviPathR")
```

Then, we can import enviPathR and another couple of optional packages used in
this tutorial.

```{r setup, message = FALSE}
library(enviPathR)
library(igraph)
library(knitr)
```

### `epLogin`: log into an enviPath personal account

It is necessary to hold an enviPath account to use the API and thus also
enviPathR. A free (for academic use) account can be created at the [enviPath
website](https://envipath.org/). The same login credentials can then be passed
to `epLogin`, which keeps you permanently logged in unless the cached package
cookies are cleared.

```{r login}
# Log into your enviPath account
epLogin(username, password)
```

After logging in once, all remaining methods work right out the box.

### `epList`: retrieve lists of objects from enviPath

enviPath divides its conceptual or conceptual items into different object
categories. The available elements for each object type ca be explored using
`epList`.

(Maybe using ariadne, show list of available object types)

```{r list-pkg}
# List available packages
pkg_df <- epList("package")

# Select reviewed packages
to_keep <- pkg_df$reviewStatus == "reviewed"
pkg_df <- pkg_df[to_keep, ]

# Remove lengthy description
pkg_df$description <- NULL

# View list of packages
kable(pkg_df)
```

```{r list-path}
# Select desired package
pkg_name <- "EAWAG-BBD"
pkg_id <- pkg_df$id[pkg_df$name == pkg_name]

# List pathways from desired package
path_df <- epList("pathway", pkg = pkg_id)

# View list of pathways
path_df |>
    head() |>
    kable()
```

Same can be done for all other supported types.

### `epGet`: fetch raw objects from enviPath

```{r get-one}
# Define reaction of interest
rxn_id <- "2b6bbcc5-77f4-4bed-92a9-731cdc978f6a"

# Fetch reaction object
rxn <- epGet("reaction", rxn_id)

# View part of reaction data
rxn |>
    head() |>
    print()
```

```{r get-many}
# Select pathways of interest
path_ids <- path_df$id[1:2]

# Fetch multiple pathways
paths <- epGet("pathway", init = path_ids, pkg = pkg_id)

# View part of pathway data
paths |>
    lapply(head) |>
    print()
```

It is also possible to extract specific properties from certain object types
with the argument `property`. These include pathway edges and nodes and compound
structures.

```{r get-property}
# Extract node property from pathway objects
nodes <- epGet("pathway", init = path_ids, pkg = pkg_id, property = "node")

# View nodes
print(nodes)
```

### `epLink`: map between different types of objects

(Maybe using ariadne, show links between object types)

```{r link-one}
# Map desired reaction to related compounds (educts and products)
rxn2cpd <- epLink("reaction", "compound", rxn_id)

# View reaction-to-compound mapping
kable(rxn2cpd)
```

```{r link-many}
# Map desired pathways to related compounds (nodes)
path2cpd <- epLink("pathway", "compound", init = path_ids, pkg = pkg_id)

# View pathway-to-compound mapping
kable(path2cpd)
```

### `epModel`: predict compound biotransformation pathways

enviPath provides a set of pre-trained models for chemical fate prediction and
half-life persistence of both known and unknown compounds given their SMILES
representation. The available models can be retrieved with `epList`.

```{r model-setting}
# List available model settings
set_df <- epList("setting")

# View list
kable(set_df)
```

Pathway prediction is performed with `epModel`, which uses the enviFormer model
by default. The output list contains two data frames representing the nodes
(compounds) and edges (reactions) of the predicted pathway, respectively.

```{r model-former}
# Define smiles of interest
smiles <- "ClC(Cl)=C(Cl)Cl"

# Perform pathway prediction with enviFormer
former_out <- epModel(smiles)

# View compounds in predicted pathway
kable(former_out$nodes)

# View reactions in predicted pathway
kable(former_out$edges)
```

Different model settings can be specified with the argument `setting`. For
example, the PEPPER model can be used which also returns half-lime estimates for
the predicted degradation steps.

```{r model-pepper}
# Set id for PEPPER model setting
set_name <- "Global Setting - ECC and App Domain - PEPPER"
set_id <- set_df$id[set_df$name == set_name]

# Perform pathway prediction with PEPPER
pepper_out <- epModel(smiles, set_id)

# View compounds in predicted pathway
kable(pepper_out$nodes)

# View reactions in predicted pathway
kable(pepper_out$edges)
```

Remarkably, the output of `epModel` can be directly converted to an igraph
object and visualised as a pathway graph. For simpliciy, the base R plot
function is used, but the ggraph package is recommended for more advanced graph
visualisation.

```{r model-plot}
# Convert model output to igraph object
path_graph <- graph_from_data_frame(
    pepper_out$edges,
    vertices = pepper_out$nodes
)

# Visualise predicted pathway
plot(path_graph)
```

## Reproducibility

### Cite enviPathR

```{r}
citation("enviPathR")
```

### Session info

```{r}
sessionInfo()
```
