---
title: "enviPathR"
author:
  - name: Giulio Benedetti
    affiliation: University of Turku
    email: giulio.benedetti@utu.fi
output:
    BiocStyle::html_document:
        fig_height: 7
        fig_width: 10
        toc: true
        toc_float: true
        toc_depth: 2
        number_sections: true
        self_contained: true
        code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('enviPathR')`"
vignette: >
  %\VignetteIndexEntry{enviPathR: access enviPath and predict pathways from R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

username <- Sys.getenv("EP_USERNAME")
password <- Sys.getenv("EP_PASSWORD")
```

## Introduction

enviPath is a biotransformation database and pathway prediction system focusing
on xenobiotic compounds that exacerbate environmental pollution. Although the
enviPath resources are accessible via an intuitive website, this approach cannot
ensure analytical efficiency and reproducibility. For this, enviPath provides
a REST API through which both data retrieval and pathway prediction are
possible. The current package, enviPathR, offers an R client to the enviPath API
and thus brings to R practitioners investigating environmental pollution and
xenobiotic biotransformation.

## Tutorial

### Installation

Bioconductor release version:

```{r install-release, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("enviPathR")
```

Beta version:

```{r install-devel, eval = FALSE}
remotes::install_github("Minotau-R/enviPathR")
```

Then, we can import enviPathR and another couple of optional packages used in
this tutorial.

```{r setup, message = FALSE}
library(enviPathR)
library(igraph)
library(knitr)
```

### `epLogin`: log into an enviPath personal account

```{r login}
# Log into your enviPath account
epLogin(username, password)
```

### `epList`: retrieve lists of objects from enviPath

```{r list-pkg}
# List available packages
pkg_df <- epList("package")

# Select reviewed packages
to_keep <- pkg_df$reviewStatus == "reviewed"
pkg_df <- pkg_df[to_keep, ]

# Remove lengthy description
pkg_df$description <- NULL

# View list of packages
kable(pkg_df)
```

```{r list-path}
# Select desired package
pkg_name <- "EAWAG-BBD"
pkg_id <- pkg_df$id[pkg_df$name == pkg_name]

# List pathways from desired package
path_df <- epList("pathway", pkg = pkg_id)

# View list of pathways
path_df |>
    head() |>
    kable()
```

Same can be done for all other supported types.

### `epGet`: fetch raw objects from enviPath

```{r get}
rxn_id <- "2b6bbcc5-77f4-4bed-92a9-731cdc978f6a"

rxn <- epGet("reaction", rxn_id)

print(rxn)
```

### Link

```{r link}
rxn2cpd <- epLink("reaction", "compound", rxn_id)

kable(rxn2cpd)
```

### `epModel`: predict compound biotransformation pathways

```{r model-setting}
# List available model settings
set_df <- epList("setting")

# View list
kable(set_df)
```

```{r model-former}
# Define smiles of interest
smiles <- "ClC(Cl)=C(Cl)Cl"

# Perform pathway prediction with enviFormer
former_out <- epModel(smiles)

kable(former_out$nodes)
kable(former_out$edges)
```

```{r model-pepper}
# Set id for PEPPER model setting
set_name <- "Global Setting - ECC and App Domain - PEPPER"
set_id <- set_df$id[set_df$name == set_name]

# Perform pathway prediction with PEPPER
pepper_out <- epModel(smiles, set_id)

kable(pepper_out$nodes)
kable(pepper_out$edges)
```

```{r model-plot}
# Convert model output to igraph object
path_graph <- graph_from_data_frame(
    pepper_out$edges,
    vertices = pepper_out$nodes
)

# Visualise predicted pathway
plot(path_graph)
```

## Reproducibility

```{r}
sessionInfo()
```

